Faster particle smoothing using Markov chain Monte Carlo sampling

### 2012-01-23 Paper/draft2_MCMCsmoother.tex
First commit: MCMC smoother, new state smoother and papers drafts.

\section{Introduction}

@@

Several problems are of interest; first, the inference of the \emph{filtering} distributions, $p(x_k|y_{1:k})$, an estimate of the current state given all previous observations; second, that of the \emph{smoothing} distribution, $p(x_{1:K}|y_{1:K})$ (where $K$ is the number of time steps), an estimate of the entire state sequence given all the observations; and third, that of the \emph{marginal smoothing} distributions, $p(x_{k}|y_{1:K})$ for each $k$. The estimation of these marginal smoothing distributions is not considered in this paper.

@@

A similar story applies to the problem of smoothing. In the linear Gaussian case, the entire state sequence, $x_{1:K}$, may be estimated in closed-form using the a Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965}. This works by modifying the ordinary Kalman filter results in a backward processing pass through the observations. For nonlinear or non-Gaussian models, it is possible again to resort to numerical methods. Now, the particles of the approximation are drawn from the smoothing distribution.

@@

Each particle is an entire realisation of the state process, with $K$ values corresponding to the state at each time step. The conventional method for generating these samples was described in \cite{Godsill2004}. In this paper a new method is presented for drawing samples from the smoothing distribution, using Markov chain Monte Carlo (MCMC). This novel procedure has computational advantages and greater flexibility over the previous methods.

In section~\ref{}, we review briefly review the basic particle filter and smoother algorithms. The new MCMC-based smoothing method is presented in section~\ref{}, with supporting simulations in section~\ref{}.

\section{Particle Filtering and Smoothing}

@@

Note that the choice $v_{k-1}^{(i)} = w_{k-1}^{(i)}$ corresponds to ordinary resampling, and other choices of $v_{k-1}^{(i)}$ represent auxiliary sampling schemes \cite{Pitt1999}. For the least computational expense, the proposal distribution with weights $v_{k-1}^{(i)} = 1/N_P$ (where $N_p$ is the number of particles) can be ``sampled'' by simply keeping the set of particles generated in the last step, with no resampling. (See \cite{Cappe2007,Doucet2009} for further discussion of resampling.)

@@

\begin{algorithm}
  \begin{algorithmic}
  	\STATE Initialise particles from prior, $x_{0}^{(i)} \sim p(x_{0})$.
  	\FOR{$k = 1 \dots K$}
  		\FOR{$i = 1 \dots N_P$}
  			\STATE Sample $\{ x_{k-1}^{(i)}, x_k^{(i)} \} \sim \sum_j v_{k-1}^{(j)} q(x_k|x_{k-1}^{(j)})$.
  			\STATE Weight $w_{k}^{(i)} \propto \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) } \times \frac{w_{k-1}^{(i)}}{v_{k-1}^{(i)} }$.
  			\STATE Discard $x_{k-1}^{(i)}$.
  		\ENDFOR
  	  \STATE Scale weights so that $\sum_i w_{k}^{(i)} = 1$.
  	\ENDFOR
  \end{algorithmic}
  \caption{Particle filter algorithm}
  \label{alg:PF}
\end{algorithm}

@@

The most basic particle smoother was presented in \cite{Kitigawa1996}, and is a trivial extension of the particle filter. Rather than marginalising the previous states from each particle, these past values are stored as well. Thus, at each step the algorithm generates an approximation to $p(x_{1:k}|y_{1:k})$. The output from the final processing step is an approximation to the desired smoothing distribution.

The weakness of the Kitigawa smoother (KS) is degeneracy. Every time the particles are resampled, those with low weights do not appear in the approximation at the next step, while those with high weights appear multiple times. The result is that many, or even all, particles will have the same values at early time steps. This effect is illustrated in figure~\ref{} in section~\ref{}. If the objective is only to produce a single point estimate then this might be sufficient, but to characterise the complete smoothing distribution the approximation must be rejeuvenated.

\begin{algorithm}
  \begin{algorithmic}
  	\STATE Initialise particles from prior, $x_{0}^{(i)} \sim p(x_{0})$.
  	\FOR{$k = 1 \dots K$}
  		\FOR{$i = 1 \dots N_P$}
  			\STATE Sample $\{ x_{1:k-1}^{(i)}, x_k^{(i)} \} \sim \sum_j v_{k-1}^{(j)} q(x_k|x_{k-1}^{(j)})$.
  			\STATE Weight $w_{k}^{(i)} \propto \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) } \times \frac{w_{k-1}^{(i)}}{v_{k-1}^{(i)} }$.
  		\ENDFOR
  	  \STATE Scale weights so that $\sum_i w_{k}^{(i)} = 1$.
  	\ENDFOR
  \end{algorithmic}
  \caption{Kitigawa smoother algorithm}
  \label{alg:PF}
\end{algorithm}

An improved particle smoothing algorithm was presented in \cite{Godsill2004}.

@@

\begin{itemize}
	\item Kitigawa smoother
	\item Doucet re-weighting smoother
	\item Godsill re-sampling smoother
	\item Complexities
	\item Algorithms
\end{itemize}

\section{New MCMC Particle Smoother}

\begin{itemize}
	\item Motivate - high complexity, low speed of other smoothers
	\item MCMC avoids need to calculate all weights at each step
	\item maths
	\item algorithm
	\item complexity
\end{itemize}

@@

\section{Simulations}

@@

\section{Conclusion}
The conclusion goes here.

@@

\appendices
\section{Proof of the First Zonklar Equation}
Appendix one text goes here.

>>>>

\section{Introduction} \label{sec:intro}

@@

@@

A similar story applies to the problem of smoothing. In the linear Gaussian case, the entire state sequence, $x_{1:K}$, may be estimated in closed-form using the a Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965}. This works by modifying the ordinary Kalman filter results in a backward processing pass through the observations. For nonlinear or non-Gaussian models, it is possible again to resort to numerical methods. Now, the particles of the approximation are drawn from the joint smoothing distribution.

@@

Each particle is an entire realisation of the state process, with $K$ values corresponding to the state at each time step. Methods for generating these samples have been described in \cite{Kitagawa1996,Godsill2004,Briers2010}. In this paper new methods are presented for drawing samples from the smoothing distribution, using Markov chain Monte Carlo (MCMC). These novel procedures have computational advantages and greater flexibility over previous methods.

In section~\ref{sec:basics}, we review briefly review the basic particle filter and existing joint smoother algorithms. The new MCMC-based smoothing methods are presented in sections~\ref{sec:mcmc_smoother} and~\ref{sec:new_state_smoother} with supporting simulations in section~\ref{sec:simulations}.

\section{Particle Filtering and Smoothing} \label{sec:basics}

@@

Note that the choice $v_{k-1}^{(i)} = w_{k-1}^{(i)}$ corresponds to ordinary resampling, and other choices of $v_{k-1}^{(i)}$ represent auxiliary sampling schemes \cite{Pitt1999}. For the least computational expense, the proposal distribution with weights $v_{k-1}^{(i)} = 1/N_F$ (where $N_F$ is the number of filter particles) can be ``sampled'' by simply keeping the set of particles generated in the last step, with no resampling. (See \cite{Cappe2007,Doucet2009} for further discussion of resampling.)

@@

The particle filter is summarised in algorithm~\ref{alg:PF}.

\begin{algorithm}
  \begin{algorithmic}
  	\STATE Initialise particles from prior, $x_{0}^{(i)} \sim p(x_{0})$.
  	\FOR{$k = 1 \dots K$}
  		\FOR{$i = 1 \dots N_F$}
  			\STATE Sample $\{ x_{k-1}^{(i)}, x_k^{(i)} \} \sim \sum_j v_{k-1}^{(j)} q(x_k|x_{k-1}^{(j)})$.
  			\STATE Weight $w_{k}^{(i)} \propto \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) } \times \frac{w_{k-1}^{(i)}}{v_{k-1}^{(i)} }$.
  			\STATE Discard $x_{k-1}^{(i)}$.
  		\ENDFOR
  	  \STATE Scale weights so that $\sum_i w_{k}^{(i)} = 1$.
  	\ENDFOR
  \end{algorithmic}
  \caption{Particle filter algorithm}
  \label{alg:PF}
\end{algorithm}

@@

The most basic particle smoother was presented in \cite{Kitagawa1996}, and is a trivial extension of the particle filter. Rather than marginalising the previous states from each particle, these past values are stored as well. Thus, at each step the algorithm generates an approximation to $p(x_{1:k}|y_{1:k})$. The output from the final processing step is an approximation to the desired smoothing distribution.

The weakness of this simple, filter-smoother (FS) is particle degeneracy. Every time the particles are resampled, those with low weights do not appear in the approximation at the next step, while those with high weights appear multiple times. The result is that many, or even all, particles will have the same values at early time steps. This effect is illustrated in figure~\ref{} in section~\ref{}. If the objective is only to produce a single point estimate then this might be sufficient, but to characterise the complete smoothing distribution the approximation must be rejeuvenated.

The filter-smoother is summarised in algorithm~\ref{alg:FS}.

\begin{algorithm}
  \begin{algorithmic}
  	\STATE Initialise particles from prior, $x_{0}^{(i)} \sim p(x_{0})$.
  	\FOR{$k = 1 \dots K$}
  		\FOR{$i = 1 \dots N_F$}
  			\STATE Sample $\{ x_{1:k-1}^{(i)}, x_k^{(i)} \} \sim \sum_j v_{k-1}^{(j)} q(x_k|x_{k-1}^{(j)})$.
  			\STATE Weight $w_{k}^{(i)} \propto \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) } \times \frac{w_{k-1}^{(i)}}{v_{k-1}^{(i)} }$.
  		\ENDFOR
  	  \STATE Scale weights so that $\sum_i w_{k}^{(i)} = 1$.
  	\ENDFOR
  \end{algorithmic}
  \caption{Filter-smoother algorithm}
  \label{alg:FS}
\end{algorithm}

An improved particle smoothing algorithm was presented in \cite{Godsill2004}. This works by constructing new particles from the smoothing distribution by sampling from the filtering distributions. This has the limitation that the particle support for the smoother is limited to that of the filter. The sampling procedure exploits the following factorisation of the joint distribution.

\begin{IEEEeqnarray}{rCl}
p(x_{1:K}|y_{1:K}) & = & p(x_K|y_{1:K}) \prod_{k=1}^{K-1} p(x_k|x_{k+1}, y_{1:K})
\end{IEEEeqnarray}

Particles are generated by sampling backwards in time from the factors of this expansion. At the $k^{th}$ step, a sample $\tilde{x}_{k+1:K} \sim p(x_{k+1:K}|y_{1:K})$ will already have been drawn. By sampling $x_k \sim p(x_k|\tilde{x}_{k+1}, y_{1:K})$ and appending $x_k$ to $\tilde{x}_{k+1:K}$, the state sequence is sequentially extended back in time. The recursion is initialised with $x_K \sim p(x_K|y_{1:K})$, i.e. a sample from the final filtering approximation.

The backwards conditional distributions may be expanded with Bayes rule.

\begin{IEEEeqnarray}{rCl}
p(x_k|\tilde{x}_{k+1}, y_{1:K}) & =       & p(x_k|\tilde{x}_{k+1}, y_{1:k}) \nonumber \\
                                & \propto & p(\tilde{x}_{k+1}|x_k) p(x_k|y_{1:k})
\end{IEEEeqnarray}

Substituting the filtering approximation into this expression yields a particle distribution which may be sampled easily.

\begin{IEEEeqnarray}{rCl}
\hat{p}(x_k|\tilde{x}_{k+1}, y_{1:K}) = \sum_i  \tilde{w}_k^{(i)} \delta_{x_{k}^{(i)}}(x_{k}) \\
\tilde{w}_k^{(i)} \propto w_k^{(i)} p(\tilde{x}_{k+1}|x_k^{(i)}) \label{eq:DBRS_weights}
\end{IEEEeqnarray}

The procedure may be repeated to produce as many samples from the smoother distribution as required. This backward-resampling smoother is summarised in algorithm~\ref{alg:DBRS}.

\begin{algorithm}
  \begin{algorithmic}
  	\STATE Run a particle filter to approximate $p(x_k|y_{1:k})$ for each $k$.
		\FOR{$i = 1 \dots N_S$}
			\STATE Sample $\tilde{x}_{K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{K}^{(j)}}(x_{K})$.
			\FOR{$k = K-1 \dots 1$}
				\FOR{$j = 1 \dots N_F$}
					\STATE Calculate weight $\tilde{w}_k^{(j)} = w_k^{(j)} p(\tilde{x}_{k+1}|x_k^{(j)})$.
				\ENDFOR
				\STATE Sample $\tilde{x}_{k}^{(i)} \sim \sum_j \tilde{w}_k^{(j)} \delta_{x_{k}^{(j)}}(x_{k})$.
			\ENDFOR
		\ENDFOR
  \end{algorithmic}
  \caption{Direct backward-resampling smoother algorithm}
  \label{alg:DBRS}
\end{algorithm}

A third method for drawing samples from the joint smoothing distribution, $p(x_{1:K}|y_{1:K})$, is included in \cite{Briers2010}. The principal topic of this paper is a two-filter smoother for estimation of the marginal smoothing distributions, $p(x_{k}|y_{1:K}$, by combination of the results from one forward and one backward particle filter. Particles from the joint distribution may then be generated by resampling from either the forward or backward filters in the manner of algorithm~\ref{alg:DBRS}.

The complexity of both the backward-resampling smoother of \cite{Godsill2004} and the two-filter variant of \cite{Briers2010} is $\mathcal{O}(N_F \times N_S \times K)$, where $N_F$ is the number of filter particles, $N_S$ the number of smoother particles, and $K$ the number of time steps. Methods for approximation of such algorithms with lower complexity are discussed in \cite{Klaas2006}. However, computational cost is often prohibitive, requiring either $N_S$ or $N_F$ to be held very low.

In the algorithms described so far, no new states are sampled during the smoothing stage; there is only a resampling of the filtering distributions. This limits the support of the smoother to the states that appear in the filtering particles. In \cite{Fearnhead2010}, an algorithm for estimation of the marginal smoothing distributions is described in which new states are sampled from a proposal, allowing full support. In addition, this algorithm achieves $\mathcal{O}( N_S \times K)$ complexity, linear in the number of particles!

In section~\ref{sec:mcmc_smoother}, we introduce an MCMC approach for drawing samples from the joint smoothing distribution with an improved computational complexity over the direct backward-resampling smoother. In section~\ref{sec:new_state_smoother}, we combine this procedure with the new-proposal method used in \cite{Fearnhead2010} to generate joint smoothing distribution particles with full support.

\section{New MCMC Particle Smoother} \label{sec:mcmc_smoother}

The bottleneck of the backward-resampling smoother (algorithm~\ref{alg:DBRS}) is the calculation of the sampling weights (equation~\ref{eq:DBRS_weights}). For each smoother particle and for each time step, a weight must be calculated for each of the $N_F$ filter particles, in order to draw a sample from the conditional distribution $\hat{p}(x_k|\tilde{x}_{k+1}, y_{1:K})$. The innovation here is to use Markov chain Monte Carlo (MCMC) to produce these samples instead of sampling directly.

MCMC algorithms are a class of numerical procedures which use a Markov chain to draw dependent samples from a probability distribution. A thorough introduction can be found in \cite{Gilks1996}. The basic Metropolis-Hastings (MH) \cite{Metropolis1953} sampler uses a series of steps in which a new state is drawn from a proposal distribution and accepted with a probability determined by the ratios of the target and proposal densities.

For the new MCMC smoother, the target distribution will be the conditional $\hat{p}(x_k|\tilde{x}_{k+1}, y_{1:K})$. If the current state of the chain is $x_k^{(m)}$, and a new sample is drawn from the proposal $x_k^{*} \sim q(x_k|\tilde{x}_{k+1}, y_{1:K})$, then it should be accepted with the following probability.

\begin{IEEEeqnarray}{rCl}
\alpha &=& \frac{ \hat{p}(x_k^{*}|\tilde{x}_{k+1}, y_{1:K}) q(x_k^{(m)}|\tilde{x}_{k+1}, y_{1:K}) }{ \hat{p}(x_k^{(m)}|\tilde{x}_{k+1}, y_{1:K}) q(x_k^{*}|\tilde{x}_{k+1}, y_{1:K}) }
\end{IEEEeqnarray}

A valid proposal distribution may be constructed using the particles of the filter approximation with any arbitrary weights.

\begin{IEEEeqnarray}{rCl}
q(x_k|\tilde{x}_{k+1}, y_{1:K}) & = & \sum_j \tilde{v}_k \delta_{x_{k}^{(j)}}(x_{k})
\end{IEEEeqnarray}

The resulting acceptance probability simplifies to the following:

\begin{IEEEeqnarray}{rCl}
\alpha &=& \frac{ w_k^{*} \tilde{v}_k^{(m)} }{  w_k^{(m)} \tilde{v}_k^{*} } \times \frac{ p(\tilde{x}_{k+1}|x_k^{*}) }{ p(\tilde{x}_{k+1}|x_k^{(m)}) }
\end{IEEEeqnarray}

Note that $w_k^{*}$ indicates the particle weight of the proposed new state, $x_k^{*}$.

The final state of each chain is used as a sample from the required conditional distribution. The complete MCMC backward-resampling smoother is summarised in algorithm~\ref{alg:MCMC-BRS}. Here we target the entire history conditional distribution, $\hat{p}(x_{1:k}|\tilde{x}_{k+1}, y_{1:K})$, which simpifies the initialisation of chains. This makes no practical difference to the algorithm.

\begin{algorithm}
  \begin{algorithmic}
  	\STATE Run a particle filter to approximate $p(x_{1:k}|y_{1:k})$ for each $k$.
		\FOR{$i = 1 \dots N_S$}
			\STATE Sample $\tilde{x}_{1:K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{1:K}^{(j)}}(x_{1:K})$.
			\FOR{$k = K-1 \dots 1$}
				\STATE $x_{1:k}^{(i)(0)} \gets \tilde{x}_{1:k}^{(i)}$.
				\FOR{$m = 1 \dots M$}
					\STATE Propose a new history, $x_{1:k}^{(i)*} \sim \sum_j \tilde{v}_k^{(j)} \delta_{x_{1:k}^{(j)}}(x_{1:k})$.
					\STATE With probability $\alpha = \frac{ w_k^{*} \tilde{v}_k^{(m)} }{  w_k^{(m)} \tilde{v}_k^{*} } \times \frac{ p(\tilde{x}_{k+1}|x_k^{*}) }{ p(\tilde{x}_{k+1}|x_k^{(m)}) }$, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)*}$. Otherwise, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)(m-1)}$.
				\ENDFOR
				\STATE $\tilde{x}_{1:k}^{(i)} \gets x_{1:k}^{(i)(M)}$
			\ENDFOR
		\ENDFOR
  \end{algorithmic}
  \caption{MCMC backward-resampling smoother algorithm}
  \label{alg:MCMC-BRS}
\end{algorithm}

The computational complexity of the MCMC-based algorithm is $\mathcal{O}(M \times N_S \times K)$, where $M$ is the (mean) number of MH steps used at each time step and smoothing particle. The advantage of the new procedure is that $M$ can often be much smaller $N_F$ and yet achieve the same error performance.

\begin{itemize}
	\item Motivate - high complexity, low speed of other smoothers
	\item MCMC avoids need to calculate all weights at each step
	\item maths
	\item algorithm
	\item complexity
	\item See direct simulation smoother as a single step of Gibbs Sampler
\end{itemize}

@@

@@

\section{Improving the smoother support} \label{sec:new_state_smoother}

\begin{itemize}
	\item Outline Fearnhead smoother
	\item Discuss extension to joint estimation (possibility of IS)
	\item MCMC method
\end{itemize}

\section{Simulations} \label{sec:simulations}

@@

\section{Conclusion} \label{sec:conclusions}
The conclusion goes here.

@@

\appendices
\section{Appendix Title}
Appendix one text goes here.
